Behavior Driven Development in Haskell

### 2012-06-01 paper.tex
Wording
Best current practices: mkl's initial version

\titlebanner{A draft paper to be submitted to Haskell Symposium 2012}
\preprintfooter{TBD: short description of paper}   % 'preprint' option specified.

@@

\authorinfo{Simon HENGEL}
           {}
           {sol@typeful.net}
%\authorinfo{Markus KLINIK}
%           {Elektrobit Automotive GmbH}
%           {markus.klinik@elektrobit.com}
\authorinfo{Kazuhiko YAMAMOTO}
           {IIJ Innovation Institute Inc.}
           {kazu@iij.ad.jp}

@@

The word ``behavior'' is associated with imperative actions.  Since
pure functions of Haskell are static, this word might be improper for
Haskell.  However, this word is already used in the Haskell
community~\cite{typeclassopedia}.  So, we use ``behavior'' instead of
defining terminology in this paper.

@@

This section show best current practices of
BDD in Haskell.

\subsection{Procedure of BDD}

First, we write a signature and leave the definition of the function undefined.

\begin{quote}
\small
\begin{verbatim}
reverse :: [a] -> [a]
reverse = undefined
\end{verbatim}
\end{quote}

\noindent Next we write user documentation in Haddock markup:

\begin{quote}
\small
\begin{alltt}
-- |
-- 'reverse' @xs@ returns the elements of @xs@
-- in reverse order. @xs@ must be finite.
--
-- >>> reverse [1,2,3]
-- [3,2,1]
--
-- prop> reverse [] == reverse []
-- prop> reverse (xs ++ ys) == reverse ys ++ \(\hookleftarrow\)
                               reverse (xs::[Int])
-- prop> reverse (reverse xs) == (xs::[Int])

reverse :: [a] -> [a]
reverse = undefined
\end{alltt}
\end{quote}

\noindent We should execute Doctest to ensure no syntax errors.
Also, we write code documentation with Hspec:

\begin{quote}
    sol: TBD. @kazu, the descriptions make no sense, can you try to
    make them more precise?
\end{quote}

\begin{quote}
\small
\begin{verbatim}
describe "reverse" $ do
  it "returns the first element in the last" $
    property $ \xs -> not (null xs) ==>
      head (reverse xs) == last (xs :: [Int])

  it "returns the last element in the first" $
    property $ \xs -> not (null xs) ==>
      last (reverse xs) == head (xs :: [Int])
\end{verbatim}
\end{quote}

\noindent Also, we should execute this Hspec module
to ensure no syntax errors.
Then we implement the function.

\begin{quote}
\small
\begin{verbatim}
reverse :: [a] -> [a]
reverse = foldl (flip (:)) []
\end{verbatim}
\end{quote}

\noindent Now we run tests, and improve our implementation, until all
tests pass.

@@

\noindent This can be used to document and automatically test
arbitrary Monoid instances.
Here is an example how to use it for lists:

@@

\begin{itemize}
    \item \emph{optional}: Bridging the gap between HUnit and QuickCheck
    \item
        Feedback for failing QuickCheck properties is not always as
        useful as it could be -> combinators that use QuickCheck's
        Result type could help
    \item
        Are more high-lever approaches to BDD, akin to Ruby's
        Cucumber, applicable to Haskell? (Observation: Cucumber's
        approach is tightly tied to OO)
\end{itemize}

@@

The {\tt test-framework} family is a TDD tool,
which currently integrates HUnit, QuickCheck, SmallCheck, and Doctest.
Automatic test discovery based on prefixes of function name
and is implemented using Template Haskell~\cite{template-haskell}.
All examples and properties must be written in a single
source file.

@@

@@

We are grateful to David Waern, Haddock's maintainer for reviewing and
accepting our patches to Haddock.
We would like to thank
Nick Smallbone for having improved QuickCheck so that
non-quantified properties are tested just once.

>>>>

% These are ignored, unless 'preprint' option specified.
\titlebanner{A draft paper to be submitted to Haskell Symposium 2012}
\preprintfooter{TBD: short description of paper}

@@

\authorinfo{Kazuhiko YAMAMOTO}
           {IIJ Innovation Institute Inc.}
           {kazu@iij.ad.jp}
\authorinfo{Simon HENGEL}
           {}
           {sol@typeful.net}
\authorinfo{Markus KLINIK}
           {Elektrobit Automotive GmbH}
           {markus.klinik@elektrobit.com}

@@

\noindent The word ``behavior'' might be associated with ``imperative
action''.  Since pure functions of Haskell are static, this word might
be improper for Haskell.  However, this word is already used in the
Haskell community~\cite{typeclassopedia}.  So, we use ``behavior''
instead of defining terminology in this paper.

@@

This section shows how behaviour driven development works in practice,
by means of the fibonacci function.  Of course this is a very simple
example, and the reader is encouraged to imagine a more sophisticated
library, where design decisions might not be as obvious as here.

We present a step-by-step refinement of a single file {\tt Fib.hs}
that contains both implementation and specs.  Each line of code is
only given once when introduced, subsequent code snippets are
incremental updates.  The reader is asked to substitute code parts as
appropriate.

\subsection{Step 1}

We start with the function's documentation, which gives a usage
example, together with the simplest possible implementation that makes
the example work.

\begin{quote}
\small
\begin{verbatim}
module Fib where
-- | Calculate Fibonacci numbers.
--
-- >>> fib 10
-- 55
fib :: Int -> Integer
fib = const 55
\end{verbatim}
\end{quote}

This first step might appear useless, but as of now we already have
made some choices with regard to the API of our library.  By giving a
usage example, we were forced to think about which arguments our
function takes.  We can now run Doctest to verify that the example
works.

\begin{quote}
\small
\begin{verbatim}
$ doctest Fib.hs
There is one test, with one single interaction.
Examples: 1  Tried: 1  Errors: 0  Failures: 0
\end{verbatim}
\end{quote}

\subsection{Step 2}

We add a specification that forces our function to actually calculate
fibonacci numbers.

\begin{quote}
\small
\begin{verbatim}
import Control.Applicative
import Test.Hspec.ShouldBe
import Test.QuickCheck

-- small non-negative numbers
newtype Small = Small Int
  deriving Show

instance Arbitrary Small where
  arbitrary = Small . (`mod` 10) <$> arbitrary

main = hspec $ do
  describe "fib" $ do
    it "calculates arbitrary Fibonacci numbers" $ do
      property $ \(Small n) ->
        fib n == fib (n + 2) - fib (n + 1)
\end{verbatim}
\end{quote}

\noindent Of course, this test now fails when we run it,

\begin{quote}
\small
\begin{verbatim}
$ runhaskell Fib.hs
fib
 - calculates arbitrary Fibonacci numbers FAILED [1]
\end{verbatim}
\end{quote}

\noindent so we add a working implementation.  Both Doctest and Hspec
are happy now.

\begin{quote}
\small
\begin{verbatim}
fib 0 = 0
fib 1 = 1
fib n = fib (n - 1) + fib (n - 2)
\end{verbatim}
\end{quote}

\subsection{Step 3}

As the seasoned Haskeller might have already noticed, our
implementation is quite inefficient, as can be proved by this
specification:

\begin{quote}
\small
\begin{verbatim}
import Control.Exception
import System.Timeout

it "is efficient" $ do
  timeout 10000
      (evaluate $ fib 32) `shouldReturn` Just 2178309
\end{verbatim}
\end{quote}

\noindent When we run it, hspec tells us

\begin{quote}
\small
\begin{verbatim}
$ runhaskell Fib.hs
fib
 - calculates arbitrary Fibonacci numbers
 - is efficient FAILED [1]
\end{verbatim}
\end{quote}

\noindent and we respond with an efficient implementation.

\begin{quote}
\small
\begin{verbatim}
import Data.List

fib :: Int -> Integer
fib n = fibs !! n
fibs = 0 : 1 : zipWith (+) fibs (tail fibs)
\end{verbatim}
\end{quote}

\subsection{Step 4}

So far, so good.  We have an implementaion of the fibonacci function
that swiftly calculates correct values for all positive integers.
But what about negative ones? Our implementation uses the list index
operator {\tt (!!)} which is a partial function that throws an
exeption for negative indexes.  Let's specify that behaviour.

\begin{quote}
\small
\begin{verbatim}
it "throws ErrorCall on negative input" $ do
  evaluate (fib (-10)) `shouldThrow` anyErrorCall
\end{verbatim}
\end{quote}

\noindent Throwing exceptions from pure code is considered bad
practice, so we guard the {\tt (!!)} operator against negative
indexes.

\begin{quote}
\small
\begin{verbatim}
fib n | n < 0     = 0
      | otherwise = fibs !! n
\end{verbatim}
\end{quote}

\noindent Now our specification fails, so we adapt it accordingly.

\begin{quote}
\small
\begin{verbatim}
it "returns 0 on negative input" $ do
  fib (-10) `shouldBe` 0
\end{verbatim}
\end{quote}

\noindent In fact, it is the case that fib returns 0 for all negative
numbers.

\begin{quote}
\small
\begin{verbatim}
it "returns 0 on negative input" $ do
  property $ \n -> n < 0 ==>
    fib n == 0
\end{verbatim}
\end{quote}

\noindent This concludes our example coding session.  We have seen how
to utilize behavior driven development to incrementally build a small
library.  The combinators {\tt shouldBe}, {\tt shouldReturn} and {\tt
shouldThrow} can be used to specify behavior of functions in an
intuitive, easy to read manner.  The final version of our library
looks like this.

\begin{quote}
\small
\begin{verbatim}
module Fib where
import Control.Applicative
import Control.Exception
import Data.List
import System.Timeout
import Test.Hspec.ShouldBe
import Test.QuickCheck

-- | Calculate Fibonacci numbers.
--
-- >>> fib 10
-- 55
fib :: Int -> Integer
fib n | n < 0     = 0
      | otherwise = fibs !! n

fibs = 0 : 1 : zipWith (+) fibs (tail fibs)

-- small non-negative numbers
newtype Small = Small Int
  deriving Show

instance Arbitrary Small where
  arbitrary = Small . (`mod` 1000) <$> arbitrary

main = hspec $ do
  describe "fib" $ do
    it "calculates arbitrary Fibonacci numbers" $ do
      property $ \(Small n) ->
        fib n == fib (n + 2) - fib (n + 1)

    it "is efficient" $ do
      timeout 10000 (evaluate $ fib 32)
          `shouldReturn` Just 2178309

    it "returns 0 on negative input" $ do
      property $ \n -> n < 0 ==>
        fib n == 0
\end{verbatim}
\end{quote}

@@

\noindent This can be used to document and test arbitrary Monoid
instances.
Here is an example how to use it for lists:

@@

\begin{itemize}
    \item
        Feedback for failing QuickCheck properties is not always as
        useful as it could be -> combinators that use QuickCheck's
        Result type could help
    \item
\end{itemize}

\begin{quote}
    TBD: add ``The Cucumber Book'' to the bibliography as {\tt
    cucumber-book}.
\end{quote}

Cucumber~\cite{cucumber-book} is a higher-level approach to BDD for Ruby, and
it is typically used in combination with RSpec.
It focuses on user stories, to assist the developer in delivering
value to the stakeholder.
There is nothing comparable to Cucumber for Haskell, yet.  And we think
it would be interesting to explore, whether it is possible to apply
Cucumber's approach to BDD to Haskell.

@@

The {\tt test-framework} family is a TDD tool, which currently
integrates HUnit, QuickCheck, SmallCheck, and Doctest.
There is support for automatic test discovery that relies on naming
conventions for test functions\footnote{%
http://hackage.haskell.org/package/test-framework-th}.  It is
implemented with Template Haskell~\cite{template-haskell}, and it only
works within a single source file.  Automatic discovery across
multiple source files is not possible with a mechanism that is based
on Template Haskell, because Template Haskell can not generate import
statements\footnote{%
See http://hackage.haskell.org/trac/ghc/ticket/1475 }.

@@

\begin{quote}
    TBD: Add "best current practice"
\end{quote}

@@

We are grateful to David Waern, Haddock's maintainer, for reviewing and
accepting our patches to Haddock.
We would like to thank
Nick Smallbone for having improved QuickCheck so that
non-quantified properties are tested just once.
